---
title: "GR5291 Advanced Data Analysis Problem Set Survival Analysis"
author: "Francis Zhang"
date: "November 22, 2024"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question

Consider the Mayo Clinic Lung Cancer Data in R package survival : data(lung) or data(cancer): including the variables

       inst:       Institution code

       time:       Survival time in days

       status:     censoring status 1=censored, 2=dead

       age:        Age in years

       sex:        Male=1 Female=2, etc.

1.Estimate and plot the survival curve for the combined Male and Female data using the following methods:

a.Kaplan-Meier

b.Fleming-Harrington

2.Estimate the median survival time, using the estimated survival curves from 1a and 1b.

3.Using a log-rank test, compare the survival distributions for Male and Female

## Solution

### Question 1

``` {r 1}
# Load necessary libraries
library(survival)
library(survminer)

# Load the lung dataset
data(lung)
head(lung)
```

#### a.Kaplan-Meier

``` {r 1.a}
# Create a survival object
surv_object <- Surv(time = lung$time, event = lung$status == 2)

# Fit the Kaplan-Meier survival model
km_fit <- survfit(surv_object ~ 1, data = lung)
summary(km_fit)

# Plot the Kaplan-Meier survival curve
ggsurvkm <- ggsurvplot(km_fit, conf.int = TRUE,
                       title = "Kaplan-Meier Survival Curve for Combined Data",
                       xlab = "Time in Days", ylab = "Survival Probability")
ggsurvkm
```

#### b.Fleming-Harrington

``` {r 1.b}
# Fit the Fleming-Harrington survival model
fh_fit <- survfit(surv_object ~ 1, type = "fh", data = lung)
summary(fh_fit)

# Plot the Fleming-Harrington survival curve
ggsurvfh <- ggsurvplot(fh_fit, conf.int = TRUE, title =
                         "Fleming-Harrington Survival Curve for Combined Data",
                       xlab = "Time in Days", ylab = "Survival Probability")
ggsurvfh
```

## Question 2

Median survival means half of the people survived at that time. According to the summary tables above, median points should be around 310.

``` {r 2}
# Kaplan-Meier Median Survival Time:
km_median <- summary(km_fit)$table["median"]
km_median
ggsurvkm$plot +
  geom_vline(xintercept = km_median, linetype = "dashed", color = "blue") +
  geom_hline(yintercept = 0.5, linetype = "dotted", color = "red") +
  annotate("text", x = km_median, y = 0.55, label =
             paste("Median:", round(km_median, 1)), color = "blue",
           hjust = -0.1)

# Fleming-Harrington Median Survival Time:
fh_median <- summary(fh_fit)$table["median"]
fh_median
ggsurvfh$plot +
  geom_vline(xintercept = fh_median, linetype = "dashed", color = "blue") +
  geom_hline(yintercept = 0.5, linetype = "dotted", color = "red") +
  annotate("text", x = fh_median, y = 0.55, label =
             paste("Median:", round(fh_median, 1)), color = "blue",
           hjust = -0.1)
```

## Question 3

The log-rank test is used to compare the survival distributions of two samples.

``` {r 3}
# Convert sex to a factor variable
lung$sex <- factor(lung$sex, levels = c(1, 2), labels = c("Male", "Female"))

# Fit the Kaplan-Meier survival model by sex
km_sex_fit <- survfit(surv_object ~ sex, data = lung)

# Plot the Kaplan-Meier survival curves by sex
ggsurvplot(km_sex_fit, conf.int = TRUE,
           title = "Kaplan-Meier Survival Curves by Sex",
           xlab = "Time in Days", ylab = "Survival Probability",
           legend.title = "Sex", legend.labs = c("Male", "Female"))

# Perform the log-rank test
log_rank_test <- survdiff(surv_object ~ sex, data = lung)
log_rank_test
```

The Kaplan-Meier survival analysis indicates that females tend to have a higher survival probability compared to males over time. The log-rank test confirms a statistically significant difference in survival distributions between males and females ( $\chi^2$ = 10.3 ,  $p$ = 0.001 ), suggesting that sex significantly impacts survival in this dataset.